---
title: "PDX HBCx-95 scRNAseq"
author: Céline Vallot
date: 21/04/2020
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries}
library(devtools)
library(dplyr)
library(DropletUtils)
library(irlba)
library(corrplot)
library(ConsensusClusterPlus)
library(geco.unsupervised)
library(scatterplot3d)
library(scater)
library(Rtsne)
library(ccRemover)
library(colorRamps)
library(geco.supervised)
library(viridis)
library(colorRamps)
library(RColorBrewer)
library(scTools)
library(edgeR)
library(ggplot2)
library(rgl)
library(RColorBrewer)
library(genefilter)
library(xtable)*
library(WriteXLS)
library(data.table)
library(stringr)
library(limma)
library(edgeR)
library(monocle3)
library(dplyr)
library(WriteXLS)

#Geco packages
library(geco.supervised)
library(geco.RNAseq)
library(geco.utils)
library(geco.visu)
library(scTools)
```


```{r paths}
#Global variables
source(file.path("../../global_var.R"))

resdir <- file.path(working_directory,"output","scRNAseq","PDX")
QCdir <- file.path(resdir,"QC")
resSUBdir <- file.path(resdir, paste0("Unsupervised")) ; if(!file.exists(resSUBdir)){dir.create(resSUBdir)}
RDatadir <- file.path(resSUBdir,"RData") ; if(!file.exists(RDatadir)){dir.create(RDatadir)}
```

```{r loads}
load(file.path(RDatadir,"HBCx95.RData"))
load(file.path(RDatadir,"pca.RData")) #if starting from prior analysis
load(file.path(RDatadir,"annot_anocol_final.RData"))#if starting from prior analysis
load(file.path(RDatadir,"umap.RData"))#if starting from prior analysis
annotationDatabases <- c("MSigDB") # "KEGG" or "GO" or "KEGG_GO" or "MSigDB"
```

We can start from already performed analysis to make figures.

### Analyzing all cells from v3 10X kits with all genes

Here, we use Monocle3 to reduce dimensions using PCA and UMAP, based on all genes. Cells are colored according to sample_id, cluster membership (louvain), expression of ribosomal genes and total counts.

### Studying cell cycle with Seurat, calculate a cell cycle score

Here we use CellCycleScoring function to annotate cells with a cell cycle phase.
By the way we perform as a comparison PCA and UMAP reduction using Seurat. The main difference with our prior analysis is that Seurat performs PCA solely on the most varying genes (n=2000) and not the entire dataset. The segragation of cells seems less efficient in 2-D space then with all genes.

```{r, echo=FALSE}
library(Seurat)
# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Create our Seurat object and complete the initalization steps
rownames(Signal) <- gene_metadata$Symbol
annot_sel <- annot[annot$kit %in% c("v3"),]

marrow <- CreateSeuratObject(counts = Signal[,annot_sel$cell_id])
marrow <- NormalizeData(marrow)
marrow <- FindVariableFeatures(marrow, selection.method = "vst")
marrow <- ScaleData(marrow, features = rownames(marrow))
marrow <- RunPCA(marrow, features = VariableFeatures(marrow), ndims.print = 1:10, nfeatures.print = 10)
DimPlot(marrow, reduction = "pca")
DimHeatmap(marrow, dims = 1:3, cells = 1000, balanced = TRUE)
ElbowPlot(marrow)
marrow <- FindNeighbors(marrow, dims = 1:20)
marrow  <- FindClusters(marrow, resolution = 0.5)
marrow <- RunUMAP(marrow, dims = 1:20)

DimPlot(marrow, reduction = "umap")

marrow <- CellCycleScoring(marrow, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)
annot_seurat <- data.frame (Seurat_cluster=paste0("C",marrow$seurat_clusters),cell_cycle=marrow$Phase)
annot_seurat$cell_id <- rownames(annot_seurat)
if(all.equal(rownames(annot_sel),rownames(annot_seurat))) annot_sel2 <- cbind(annot_sel,annot_seurat) else print("ERROR")

FeaturePlot(marrow, features = c("KDM6B", "MGP","IGFBP3","KRT6A","KRT14"))
annotCol <- c("sample_id","total_features","Seurat_cluster","cell_cycle")
pcaText <- FALSE
annotText <- "sample_id"
hcText <- "sample_id"  
rm(marrow); gc()

```

###Study cells with PCA, UMAP & hierarchical clustering

```{r, echo=FALSE}
if(!(file.exists(file.path(RDatadir,"NormCounts.RData")) & file.exists(file.path(RDatadir,"LogCounts.RData")))){
    rownames(Signal) <- gene_metadata$Symbol
    rownames(annot_sel2) <- annot_sel2$cell_id
    rownames(gene_metadata) <- gene_metadata$Symbol
    cds <- new_cell_data_set(Signal[,rownames(annot_sel2)],
                             cell_metadata = annot_sel2,
                             gene_metadata  = gene_metadata)
    rm(Signal);gc()
    cds <- preprocess_cds(cds,method='PCA', norm_method='size_only',num_dim=50)
    
    cds <- reduce_dimension(cds, reduction_method = 'UMAP')
    cds <-  cluster_cells(cds)
    
    annot_sel2$louvain_cluster <- cds@clusters[[1]]$clusters
    annot_sel2$louvain_cluster <- paste0("C",annot_sel2$louvain_partition)
    
    annot_sel2$louvain_partition <- cds@clusters[[1]]$partition
    annot_sel2$louvain_partition <- paste0("C",annot_sel2$louvain_partition)
    
    umap_res <- cds@reducedDims$UMAP # with R3.6.2 and higher for SummarizedExperiment, otherwise cds@reducedDims[[2]]
    pca_object <- cds@reducedDims$PCA
    
    save(umap_res,file=file.path(RDatadir,"umap.RData"))
    save(pca_object,file=file.path(RDatadir,"pca.RData"))
    
    NormCounts <- t(t(exprs(cds)) /  pData(cds)[, 'Size_Factor']) #keep format sparse matrix, need to have row and colnames separate
    LogCounts <- log(NormCounts+1,2)
    
    save(NormCounts,file=file.path(RDatadir,"NormCounts.RData"))
    save(LogCounts,file=file.path(RDatadir,"LogCounts.RData"))
    rm(NormCounts); gc()
    rm(cds); gc()
}
```


###Perform hierarchical clustering
```{r , out.width="30%",include=FALSE}
mati <- t(pca_object)
hc_PCA <- hclust(as.dist(1 - cor(mati)), method = "ward.D")
mat.so.cor <- mati[,hc_PCA$order]

nclust <- 5
annot_sel2$clustCol5 <- paste("C",match(cutree(hc_PCA,nclust),unique(cutree(hc_PCA,nclust)[hc_PCA$order])),sep="")
nclust <- 6
annot_sel2$clustCol6 <- paste("C",match(cutree(hc_PCA,nclust),unique(cutree(hc_PCA,nclust)[hc_PCA$order])),sep="")
nclust <- 4
annot_sel2$clustCol4 <- paste("C",match(cutree(hc_PCA,nclust),unique(cutree(hc_PCA,nclust)[hc_PCA$order])),sep="")
```

```{r, echo=FALSE}
load(file.path(RDatadir,"LogCounts.RData"))
load(file.path(RDatadir,"annot_anocol_final.RData"))

diff_genes <-  c("KRT17", "INHBB", "KRT5","KRT8","NNMT","TAGLN","INHBA","KRT6A","KRT6B","KRT16","KRT14","KLK10","KLK5","KLK7")
diff_genes_figures <-  c("BMP6","CDKN2B","TGFBR2","TGFBR3","TGFB2","INHBA","INHBB","SMAD2","TGFB1","FOSL1","NNMT","KRT14","KLK10","KLK5","TAGLN","NNMT","ELN","TGFBR3","MIF")
hallmark_EMT <- c("FN1","TAGLN","NNMT","ELN","INHBA","MYL9","FBLN2","TGFBR3","CRLF1")

KEGG_TGFB_FC1 <- c("INHBB","INHBA","CREBBP","ROCK1","ID4","RBL2","SMAD4","SMAD1","SMAD2","MAPK1","SMURF1","EP300","TGFB2","CDKN2B","PPP2R1A","SMAD5","FST","TGFBR2","SMAD7","BMPR2","BMPR1B","ACVR2B","ACVR2A","SP1","BMP6","THBS1")
TGF_pathway <- c("TGFBR3","TGFBR2","TGFB2","INHBA","INHBB")

KEGG_TGFB_FC2 <-  c("CDKN2B","ID4","BMP6","INHBA","INHBB")

hallmark_TNFA <- c("CXCL2","ATF3","NFKBIA","TNFAIP3","IER3","CCL20","MAFF","BIRC3","PLAUR","ZFP36","PPP1R15A","NR4A2","MAP3K8","IL6","FOSL1","NR4A1","REL","NR4A3","GEM","GADD45A","PLK2","BHLHE40","NFIL3","TRIB1","TIPARP","INHBA","SGK1","FOSL2","CDKN1A","CYR61","IFIT2","IRF1","FOS","KLF4","LAMB3","CCNL1","CLCF1","IRS2","DUSP4","MXD1")

hallmark_IFN <- c("MX1","BST2","OASL","IFI44L","IFI27","PTGS2","HLA-B")
KRT_signature <- c("KRT6A","KRT6B","KRT79","KRT14","KRT80","KRT23","KRT16","KRT15")
hallmark_apoptosis <- c("BIRC3","BIK","CDKN1A","DDIT3","IL18","TNFRSF12A","HSPB1","ATF3","HMOX1","IER3","TGFB2","TXNIP","ANXA1")


NRG1_signalling <- c("NFIL3","ID1","TIPARP","MIR22HG","AVPI1","ATF3","CLCF1","PPP1R15A","GADD45A","CYP24A1","ZFP36","USP36","CYR61","TRIB1","MAP3K8","BHLHE40","ZNF331","GEM","TNFRSF11B","PLAUR","MAFF","NR4A2","DLX2","IER3","SGK1","FOS","REL","DUSP4","CTH","FOSL1","IL6R","SPRR1B","CCNL1","IRS2","NR4A3","NR4A1","DDIT3")

MYB_targets <- c("HSPB1","MAP1LC3B","S100A7","CDKN1A","HMOX1","SERPINB2","RRAD","ANXA1","KRT17","ANXA3","MYL12A","TUBA4A","KRT16","IL8","ARID5B","IGFBP7","KRT14","AHNAK2","KRT6A","PURA","NNMT")

SMID_LUMINAL_DN <- c("ALDH1A3","NNMT","ID4","KYNU","KRT6A","NOV","KRT23","DSC3","TNFRSF11B","CLIC3","IL8","ANXA1","ANXA3","BIRC3","TMEM45A","KLK7","DSC2","KRT17","ADM","KLK6","KLK10","ID1","KRT16","S100A7","KRT14","TAGLN","PERP")

annotCol <- c("sample_id","total_features","cell_cycle","rRNA","clustCol4","clustCol5","clustCol6","louvain_partition",
              "louvain_cluster","Seurat_cluster",
              "MKI67","KEGG_TGFB_FC1","KEGG_TGFB_FC2",diff_genes_figures,"TGF_pathway","hallmark_EMT","MYB_targets",
              "hallmark_TNFA")

pcaText <- FALSE
annotText <- "sample_id"

genes <- c("KEGG_TGFB_FC1","KEGG_TGFB_FC2","TGF_pathway","hallmark_EMT","hallmark_apoptosis","hallmark_TNFA","NRG1_signalling","MYB_targets","SMID_LUMINAL_DN")

for(i in 1:length(genes)) {
genelist <- get(genes[i])
annot_sel2$test <- apply(LogCounts[genelist,],2,mean)
colnames(annot_sel2)[dim(annot_sel2)[2]] <- genes[i]
}
for(i in annotCol){
    if(i %in% rownames(LogCounts)) annot_sel2[,i] <- LogCounts[which(rownames(LogCounts)==i),]; gc()
}

load(file.path(working_directory,"output","scRNAseq","common_overexpressed_genes_pers_vs_unt.RData"))

for(i in common_overexpressed_genes){
  annot_sel2[,i] = LogCounts[i,annot_sel2$cell_id]
}
annotCol = unique(c(annotCol, common_overexpressed_genes))

ribo <- grepl("^RP[SL]",gene_metadata$Symbol)
annot_sel2$rRNA <- apply(LogCounts[ribo,],2,mean)

library(geco.unsupervised)
anocol2 <- geco.annotToCol4(annotS=annot_sel2[,annotCol],annotT=annot_sel2,plotLegend=T,plotLegendFile=file.path(resSUBdir,"Annotation_legends.pdf"),scale_q = "inferno")


#Common palette for all MM468 experiments
control <- c("#E0E0E0","#BDBDBD","#757575")
persister_color <- c("#DCEDC8","#9CCC65","#4CAF50","#009688")
treated_UNC_color <- c("#2196F3", "#4DD0E1" )
treated_GSKJ4_color <- c("#C2185B", "#EC407A" )
res_color <- c("#FFEB3B","#FFC107","#FF9800","#FF5722")
color_PDX <- c(control[c(1,2)],persister_color[c(1,3)],res_color[c(2)],"#ffcbcb")

corres <- data.frame(PDX=c("HBCx95_v3_UNT","HBCx95_m9_50_REC","HBCx95_persister_2souris","HBCx95_persister_6souris","HBCx95_m30_CAPAR","HBCx95_m6_CAPAS"),color=color_PDX)
anocol2[,"sample_id"] <- as.character(corres$color[match(annot_sel2$sample_id,corres$PDX)])

corres_cell_cycle <- data.frame(phase=c("G1","G2M","S"),color=c("#e896aaff","#553fc2ff","#5d6c7dff"))
indice <- which(annotCol=="cell_cycle")
anocol2[,indice] <- as.character(corres_cell_cycle$color[match(annot_sel2$cell_cycle,corres_cell_cycle$phase)])

corres_cluster <- data.frame(phase=c("C1","C2","C3","C4","C5"),color=c("#b9cced","#6a8caf","#438a5e","#e2979c","#ff9c71"))
indice <- which(annotCol=="clustCol5")
anocol2[,indice] <- as.character(corres_cluster$color[match(annot_sel2$clustCol5,corres_cluster$phase)])
corres_cluster <- data.frame(phase=c("C1","C2","C3","C4","C5"),color=c("#438a5e","#6a8caf","#ff9c71","#e2979c","#b9cced"))
indice <- which(annotCol=="louvain_partition")
anocol2[,indice] <- as.character(corres_cluster$color[match(annot_sel2$louvain_partition,corres_cluster$phase)])

png(file.path(resSUBdir,"Legend_sample_scRNAseq.png"), height=2000,width=1500,res=300)
par(mfrow=c(5,2))
barplot(rep(1,6),col=corres$color, cex.names  =0.8,horiz=F,names.arg=corres$PDX,las=2)
dev.off()

png(file.path(resSUBdir,"Legend_cluster_scRNAseq.png"), height=2000,width=1500,res=300)
par(mfrow=c(5,2))
barplot(rep(1,5),col=corres_cluster$color, cex.names  =0.8,horiz=F,names.arg=c("C1","C2","C3","C4","C5"),las=2)
dev.off()


png(file.path(resSUBdir,"Boxplot_KEGG_TGFB_FC2.png"),width=1500,height=1500,res=300)
boxplot(annot_sel2$KEGG_TGFB_FC2~annot_sel2$sample_id,las=2)
dev.off()
png(file.path(resSUBdir,"Boxplot_KEGG_TGFB_FC1.png"),width=1500,height=1500,res=300)
boxplot(annot_sel2$KEGG_TGFB_FC1~annot_sel2$sample_id,las=2)
dev.off()
png(file.path(resSUBdir,"Boxplot_INHBA.png"),width=1500,height=1500,res=300)
boxplot(annot_sel2$INHBA~annot_sel2$sample_id,las=2)
dev.off()

png(file.path(resSUBdir,"Boxplot_rRNA.png"),width=1500,height=1500,res=300)
boxplot(annot_sel2$rRNA~annot_sel2$sample_id,las=2)
dev.off()
save(anocol2,annot_sel2,file=file.path(RDatadir,"annot_anocol_final.RData")) 
```

###Generate figures: UMAP and clustering with correlation heatmap

```{r, out.width="30%", echo=FALSE}
load(file.path(RDatadir,"umap.RData"))
for(j in 1:ncol(anocol2))
    {
       if(class(annot_sel2[1,colnames(anocol2)[j]])=="numeric"){
    plot((umap_res), col=alpha(anocol2[,j],0.3),pch=20,cex=0.6,
         main=paste0(colnames(anocol2)[j]," min=",round(min(annot_sel2[,colnames(anocol2)[j]]),digits=3)," max=",round(max(annot_sel2[,colnames(anocol2)[j]]),digits=3)),
         xlab="component 1",ylab="component 2")} else {
  plot((umap_res), col=alpha(anocol2[,j],0.3),pch=20,cex=0.6,
         main=paste0(colnames(anocol2)[j]),
         xlab="component 1",ylab="component 2")
         }
  
}

for(j in 1:ncol(anocol2))
    {
    png(file.path(resSUBdir,paste0("UMAP_",colnames(anocol2)[j],".png")), height=1350,width=1200,res=300)
  if(class(annot_sel2[1,colnames(anocol2)[j]])=="numeric"){
    plot((umap_res), col=alpha(anocol2[,j],0.3),pch=20,cex=0.6,
         main=paste0(colnames(anocol2)[j]," min=",round(min(annot_sel2[,colnames(anocol2)[j]]),digits=3)," max=",round(max(annot_sel2[,colnames(anocol2)[j]]),digits=3)),
         xlab="component 1",ylab="component 2")} else {
             plot((umap_res), col=alpha(anocol2[,j],0.3),pch=20,cex=0.6,
         main=paste0(colnames(anocol2)[j]),
         xlab="component 1",ylab="component 2")
             if(colnames(anocol2)[j]=="sample_id"){
                 #Plot persister with add plot since there is too low cell number
                 pers = which(annot_sel2$sample_id=="HBCx95_persister_6souris")
                 points(umap_res[pers,], col=alpha(anocol2[pers,j],0.5),pch=20,cex=0.6)
                 
             }
         }
  
  
    dev.off()
    }

# Subsample plot
library(ChromSCape)
dim(annot_sel2)
scExp = create_scExp(LogCounts,annot_sel2[,1:4])
gc()
scExp = subsample_scExp(scExp,n_cells = 500)
gc()
umap_res_sub = umap_res[scExp$cell_id,]
anocol2_sub = anocol2[scExp$cell_id,]
annot_sel2_sub = annot_sel2[scExp$cell_id,]
LogCounts_sub = LogCounts[,scExp$cell_id]
gc()
save(umap_res_sub,anocol2_sub,annot_sel2_sub,LogCounts_sub, file = file.path(RDatadir,"all_sub_500.RData"))
resSUBdir_sub = file.path(resSUBdir,"sub500"); if(!dir.exists(resSUBdir_sub)) dir.create(resSUBdir_sub)
for(j in 1:ncol(anocol2_sub))
    {
    png(file.path(resSUBdir_sub,paste0("UMAP_sub500_",colnames(anocol2_sub)[j],".png")), height=1350,width=1200,res=300)
  if(class(annot_sel2_sub[1,colnames(anocol2_sub)[j]])=="numeric"){
    plot((umap_res_sub), col=alpha(anocol2_sub[,j],0.3),pch=20,cex=0.6,
         main=paste0(colnames(anocol2_sub)[j]," min=",round(min(annot_sel2_sub[,colnames(anocol2_sub)[j]]),digits=3)," max=",round(max(annot_sel2_sub[,colnames(anocol2_sub)[j]]),digits=3)),
         xlab="component 1",ylab="component 2")} else {
  plot((umap_res_sub), col=alpha(anocol2_sub[,j],0.3),pch=20,cex=0.6,
         main=paste0(colnames(anocol2_sub)[j]),
         xlab="component 1",ylab="component 2")
         }
  
  
    dev.off()
}

png(file.path(resSUBdir,"Clustering_correlation_matrix_PCA_all_annot.png"), height=1500,width=1500,res=300)
geco.hclustAnnotHeatmapPlot(x=cor(mat.so.cor),
                            hc=hc_PCA,
                            hmColors=corColors,
                            anocol=anocol2[hc_PCA$order,],#[,ncol(cc.col):1]
                            xpos=c(0.15,0.9,0.164,0.885),
                            ypos=c(0.1,0.5,0.5,0.6,0.62,0.95),
                            dendro.cex=0.1,
                            xlab.cex=0.2,
                            hmRowNames=FALSE,
                            hmRowNames.cex=0.01
)
dev.off()


ClusterMembership <- as.data.frame.matrix(table(annot_sel2$sample_id,annot_sel2$louvain_partition))
ClusterMembership$pvalue <- 0
ClusterMembership$sample_id <- rownames(ClusterMembership)
for(i in c(1:(dim(ClusterMembership)[1]))){
 test <- chisq.test(ClusterMembership[c(i,6),c(1:5)]) 
  ClusterMembership$pvalue[i] <- test$p.value
}

WriteXLS(ClusterMembership,file.path(resSUBdir,"ClusterMembership.xls"),row.names = T)
ClusterMembership$sample_id <- rownames(ClusterMembership)
ClusterMembership_b <- ClusterMembership[,-6] %>% pivot_longer(cols=c("C1","C2","C3","C4","C5"),names_to = "cluster",values_to = "freq") %>% dplyr::mutate(cluster=factor(cluster, levels =c("C3","C1","C2","C4","C5"))) %>% dplyr::mutate(sample_id=factor(sample_id, levels =c(as.character(corres$PDX[c(1,3,4,2,5,6)])))) %>% dplyr::group_by(sample_id) %>% dplyr::mutate(freq_prop=freq/sum(freq))
png(file.path(resSUBdir,"ClusterMembership.png"),width=1000,height=1200,res=300)
ggplot(ClusterMembership_b) +geom_col(aes(y=freq_prop,x=sample_id,fill=cluster),width = 0.75) + scale_fill_manual(values=as.character(corres_cluster$color[c(3,1,2,4,5)])) + theme_classic() +theme(axis.text.x=element_blank(),axis.title = element_blank(), legend.position = "None")
dev.off()
gc()
```

```{r , out.width="30%",include=FALSE}
# Sample intracorrelation

control <- c("#E0E0E0","#BDBDBD","#757575")
persister_color <- c("#DCEDC8","#9CCC65","#4CAF50","#009688")
treated_UNC_color <- c("#2196F3", "#4DD0E1" )
treated_GSKJ4_color <- c("#C2185B", "#EC407A" )
res_color <- c("#FFEB3B","#FFC107","#FF9800","#FF5722")
color_PDX <- c(control[c(1,2)],persister_color[c(1,3)],res_color[c(2)],"#ffcbcb")

corres <- data.frame(PDX=c("HBCx95_v3_UNT","HBCx95_m9_50_REC","HBCx95_persister_2souris","HBCx95_persister_6souris","HBCx95_m30_CAPAR","HBCx95_m6_CAPAS"),color=color_PDX)

intra_corr=data.frame()
for(i in unique(annot_sel2$sample_id)){
    cells = annot_sel2$cell_id[which(annot_sel2$sample_id==i)]
    tmp = cor_mat[cells,cells]
    tab = data.frame("sample_id" = rep(i,ncol(tmp)), "intra_corr" = colMeans(tmp))
    intra_corr=rbind(intra_corr,tab)
}

intra_corr$sample_id = factor(intra_corr$sample_id, levels =c("HBCx95_v3_UNT","HBCx95_m9_50_REC","HBCx95_persister_6souris","HBCx95_persister_2souris","HBCx95_m6_CAPAS","HBCx95_m30_CAPAR"))
png(file.path(resSUBdir,paste0("Sample_intracorrelation_violin.png")),height=1000,width=1000,res=300)
ggplot(intra_corr,aes(x=sample_id,y=intra_corr, fill=sample_id)) + 
  geom_violin(alpha=0.8) + theme_classic() + scale_fill_manual(values=as.character(corres$color)[c(1,2,4,3,6,5)]) +
  stat_summary(fun=median, geom="point", size=2, color="black") + theme(axis.text.x = element_text(angle=90))
dev.off()
```
###Study cell cycle in every sample

```{r , out.width="30%",include=FALSE}
#Association of cell cycle to samples
library(tidyverse)
CellCycle <- as.data.frame.matrix(table(annot_sel2$sample_id,annot_sel2$cell_cycle))
CellCycle$pvalue <- 0
for(i in c(1:(dim(CellCycle)[1]))){
 test <- chisq.test(CellCycle[c(i,6),c(1:3)]) 
  CellCycle$pvalue[i] <- test$p.value
}

WriteXLS(CellCycle,file.path(resSUBdir,"CellCycle_samples_vs_UNT.xls"),row.names = T)
CellCycle$sample <- rownames(CellCycle)
CellCycle_b <- CellCycle[,-4] %>% pivot_longer(cols=c("G1","G2M","S"),names_to = "cycle",values_to = "freq") %>% dplyr::mutate(cycle=factor(cycle, levels =c("G1","G2M","S"))) %>% dplyr::group_by(sample) %>% dplyr::mutate(freq_prop=freq/sum(freq))
png(file.path(resSUBdir,"CellCycle.png"),width=1000,height=1000,res=300)
ggplot(CellCycle_b) +geom_col(aes(y=freq_prop,x=sample,fill=cycle)) + 
    scale_fill_manual(values=as.character(corres_cell_cycle$color)) + 
    theme_classic() +theme(axis.text.x=element_text(angle=90, hjust=1)) +
    theme(text=element_blank()) # if no legend
dev.off()

```

## Run differential analysis

```{r , out.width="30%",include=FALSE}
# save(RawCounts,feature,file="Unsupervised/RData/RawCounts_features.RData")

resdir_supervised=file.path(resdir,"Supervised"); if(!dir.exists(resdir_supervised)) dir.create(resdir_supervised)
tabdir <- file.path(resdir_supervised,"Tables");if(!file.exists(tabdir)){dir.create(tabdir)}
plotdir <- file.path(resdir_supervised,"Plots");if(!file.exists(plotdir)){dir.create(plotdir)}
EnrichDir <- file.path(tabdir,"Gene_set_analysis");if(!file.exists(EnrichDir)){dir.create(EnrichDir)}
RDataSupdir <-  file.path(resdir_supervised,"RData");if(!file.exists(RDataSupdir)){dir.create(RDataSupdir)}

load("Unsupervised/RData/RawCounts_features.RData")
load(file.path(RDatadir,"all_sub_500.RData"))
RawCounts_sub = RawCounts[,annot_sel2_sub$cell_id]
gc()

metadata = annot_sel2_sub
##raw counts for edgeR
# RawCounts <- Signal[,metadata$cell_id]
# feature <- gene_metadata

#Get the gene list of Persister cells (6souris) overexpressed
mygps <- list(

'persister_vs_UNT'=metadata[which(metadata$sample_id %in% c("HBCx95_persister_6souris","HBCx95_persister_2souris") ),"cell_id"],
'persister_vs_all'=metadata[which(metadata$sample_id %in% c("HBCx95_persister_6souris","HBCx95_persister_2souris") ),"cell_id"],
'persister_6_vs_UNT'=metadata[which(metadata$sample_id %in% c("HBCx95_persister_6souris") ),"cell_id"]
)
sapply(mygps,length)

myrefs <- list(
    'UNT'=metadata[which(metadata$sample_id %in% c("HBCx95_v3_UNT") ),"cell_id"],
    'all'=metadata[which(!metadata$sample_id %in% c("HBCx95_persister_6souris","HBCx95_persister_2souris")  ),"cell_id"],
    'UNT'=metadata[which(metadata$sample_id %in% c("HBCx95_v3_UNT") ),"cell_id"]
    
)
sapply(myrefs,length)

refs <- names(myrefs) 
groups <- names(mygps) 

FC_threshold <- 2
Signif_threshold <- 0.01
algoType <- c("edgeR") ## Limma or DESeq or a vector of both or edgeR for single-cell
useBlockFactor <- FALSE ## TRUE or FALSE
#BlockFactor <- "Sex"

my.res <- geco.CompareedgeRGLM (dataMat=RawCounts_sub,
                                  annot=metadata,
                                  ref=myrefs,
                                  groups=mygps,
                                  featureTab=feature
  )
  
  
under_res = my.res %>% dplyr::filter(log2FC.persister_6_vs_UNT < -FC_threshold & qval.persister_6_vs_UNT < Signif_threshold) %>% 
    dplyr::arrange(qval.persister_6_vs_UNT) %>% select(Symbol,log2FC.persister_6_vs_UNT,qval.persister_6_vs_UNT)
  over_res = my.res %>% dplyr::filter(log2FC.persister_6_vs_UNT > FC_threshold & qval.persister_6_vs_UNT < Signif_threshold) %>% 
    dplyr::arrange(qval.persister_6_vs_UNT) %>% dplyr::select(Symbol,log2FC.persister_6_vs_UNT,qval.persister_6_vs_UNT)
  
  WriteXLS(c("over_res","under_res","my.res"), ExcelFileName =   file.path(tabdir,"Differential_analysis_Limma.xlsx"), SheetNames = c(paste0("Over_",FC_threshold,"_",Signif_threshold,"_n",nrow(over_res)),paste0("Under_-",FC_threshold,"_",Signif_threshold,"_n",nrow(under_res)),"All"), perl = "perl", verbose = FALSE, row.names = FALSE, col.names = TRUE, AdjWidth = T, AutoFilter = TRUE, BoldHeaderRow = TRUE, na = "", FreezeRow = 1, FreezeCol = 1)
  
  
  
  summaryTab <- geco.summaryCompareedgeR(restab=my.res,
                                         ref=myrefs,
                                         groups=mygps,
                                         qval.th=Signif_threshold,
                                         fc.th=FC_threshold,
                                         plotdir=plotdir
  )
  
  summaryTab <- data.frame("."=row.names(summaryTab), summaryTab, check.names=FALSE)		
  write.table(summaryTab, file.path(tabdir,"Number_differentially_expressed_genes_per_group.csv"),row.names=F,quote=F,sep=";"  )
  print("summaryedgeRCompare Done")
```

###Generate heatmaps with gene lists 

```{r , out.width="30%",include=FALSE}
#import my.res et gene lists from diff analysis, logCounts, take top 10
MSigDBFile1 <- "/media/pprompsy/LaCie/InstitutCurie/Documents/GitLab/ChromSCape_devel/data/hg38.MSIG.gs.rda" 
MSigDBFile2 <- "/media/pprompsy/LaCie/InstitutCurie/Documents/GitLab/ChromSCape_devel/data/hg38.MSIG.ls.rda"
load(file=file.path(RDataSupdir,paste("Supervised_res_object_edgeR.Rdata",sep="")))
load(MSigDBFile1)
load(MSigDBFile2)
MSIG.ls = hg38.MSIG.ls
MSIG.gs = hg38.MSIG.gs

rownames(my.res) <- my.res$Gene

annotbase <- "MSigDB" 
database <- MSIG.ls ##MSigDB

Overexpressed  <- Underexpressed <- data.frame()

reflist <- unique(my.res$Symbol);length(reflist)

for(i in 1:length(groups)) 	{
    gp <- groups[i]
    if (length(refs)>1) {ref <- refs[i]} else {ref <- refs[1]}
    print(paste0("Processing ",gp, " vs ", ref, " _ ",annotbase ))
    
    signific <- which(my.res[,paste("qval",gp,sep=".")] <= Signif_threshold & abs(my.res[,paste("log2FC",gp,sep=".")]) > FC_threshold)
    over <- which(my.res[,paste("qval",gp,sep=".")] <= Signif_threshold & my.res[,paste("log2FC",gp,sep=".")] > FC_threshold)
    under <- which(my.res[,paste("qval",gp,sep=".")] <= Signif_threshold & my.res[,paste("log2FC",gp,sep=".")] < -FC_threshold)
    print(paste0("significant = ", length(signific))) ; print(paste0("over = ", length(over))) ; print(paste0("under = ", length(under)))
    
    
    if(length(over)){
        enrich.test <- geco.enrichmentTest(gene.sets=database,mylist=my.res$Symbol[over],possibleIds=reflist)
        enrich.test <- data.frame(Gene_set_name=rownames(enrich.test), enrich.test, check.names=FALSE)
        enrich.test <- merge( subset(MSIG.gs, select=-Genes), enrich.test, by.x="Gene.Set", by.y="Gene_set_name", all.y=TRUE, sort=FALSE ) ## Get class of gene set
        
        enrich.test <- enrich.test[order(enrich.test$`p-value`),]
        ind <- which(enrich.test$`q-value`<= 0.1);if(!length(ind)){ind <- 1:20}
        Overexpressed  <- enrich.test[ind,]
      png(file.path(EnrichDir,paste0(groups[i],"_vs_",ref,"_qval",Signif_threshold,"_logFC",FC_threshold,"_enriched_peaks_enrichment.png")), height=1000,width=1000,res=150)
      par(mar = c(13, 4, 2, 2) + 1)
      barplot(-log10(enrich.test$`q-value`[1:10]),las=2,names.arg = enrich.test$Gene.Set[1:10],cex.names = 0.5,ylab="-log10 q-value",main=paste0("Top 10 gene list of ",length(ind)," lists"))
      dev.off()
        }
    
    if(length(under)){
        enrich.test <- geco.enrichmentTest(gene.sets=database,mylist=my.res$Symbol[under],possibleIds=reflist)
        enrich.test <- data.frame(Gene_set_name=rownames(enrich.test), enrich.test, check.names=FALSE)
        enrich.test <- merge( subset(MSIG.gs, select=-Genes), enrich.test, by.x="Gene.Set", by.y="Gene_set_name", all.y=TRUE, sort=FALSE ) ## Get class of gene set
        
        enrich.test <- enrich.test[order(enrich.test$`p-value`),]
        ind <- which(enrich.test$`q-value`<= 0.1);if(!length(ind)){ind <- 1:20}
        Underexpressed <- enrich.test[ind,]		}
    
    WriteXLS(c("Overexpressed", "Underexpressed"), ExcelFileName = file.path(EnrichDir,paste0("Enrichment_test_",gp,"_vs_",ref,"_",annotbase,".xlsx")), SheetNames = c( paste0("Overexp_in_", gp), paste0("Underexp_in_", gp) ), perl = "perl", verbose = FALSE, row.names = FALSE, col.names = TRUE, AdjWidth = TRUE, AutoFilter = TRUE, BoldHeaderRow = TRUE, na = "", FreezeRow = 1, FreezeCol = 1)
} 		
```

### Heatmap for a gene list of interest
```{r , out.width="30%",include=FALSE}
mat <- LogCounts[rownames(LogCounts) %in% KEGG_TGFB_FC2,]
tmat <- t(mat)

distHC <- c("distPearson","distCosine","euclidean","maximum","manhattan","canberra","binary","minkowski")[3]
methHC <- c("ward","ward.D","ward.D2","single","complete","average")[3]
if(distHC=="distPearson")	d <- distPearson(tmat)
if(distHC=="distCosine")	d <- distCosine(tmat)
if(distHC %in% c("euclidean","maximum","manhattan","canberra","binary","minkowski"))	
d <- dist(tmat,distHC)
hc <- hclust(d,method=methHC)

mat.so <- as.matrix(mat[,hc$order])

if(chRangeHM){
  for(i in 1:nrow(mat.so)){
    mat.so[i,] <- geco.changeRange(mat.so[i,],newmin=0,newmax=1)
  }
}

png(file.path(resSUBdir,paste0("Clustering_TGFB_","_",distHC,"_",methHC,".png")), height=4000,width=4000,res=300)

geco.hclustAnnotHeatmapPlot(x=(mat.so),
                            hc=hc,
                            hmColors=hmColors,
                            anocol=anocol2[hc$order,c(1:5)],#[,ncol(cc.col):1]
                            xpos=c(0.15,0.9,0.164,0.885),
                            ypos=c(0.1,0.5,0.5,0.6,0.62,0.95),
                            dendro.cex=0.01,
                            xlab.cex=1,
                            hmRowNames=TRUE,
                            hmRowNames.cex=0.5
)
dev.off()

```


###Study the top gene lists of persister cells
```{r , out.width="30%",include=FALSE}

library(gplots)

load(file=file.path(RDatadir,"Enrichment_Overexpressed_persister_6_vs_UNT.RData"))
load(file=file.path(RDatadir,"LogCounts.RData"))
load(file=file.path(RDatadir,"RawCounts_features.RData"))
gene_metadata = feature
  how_many_top <- 20
		significPathway <- Overexpressed[Overexpressed$Class %in% c("c2_curated","c6_oncogenic","hallmark"),]

	significPathway <- significPathway[1:how_many_top,]
	significPathway_breast <- 	significPathway[grep(pattern = "BREAST", significPathway$Gene.Set),]

			
	#Calculate average expression for every cells on genes of each top pathways
	pathway_mat <- matrix(0,nrow=how_many_top,ncol=length(annot_sel2$sample_id))		
	pathway_names <- c()		
	for(i in 1:how_many_top){
	  gene_list <- unlist(strsplit(significPathway$Deregulated_genes[i],";"))
	  mat <- LogCounts[gene_metadata$Symbol %in% gene_list,]
	  pathway_mat[i,] <- apply(mat,2,mean) 
	  pathway_names[i] <-significPathway$Gene.Set[i]
	}		
  
row.names(pathway_mat) <- pathway_names
tmat <- t(pathway_mat)

selected_cells = annot_sel2$cell_id[which(annot_sel2$louvain_partition %in% c("C1","C5"))]
rownames(tmat) = colnames(LogCounts)
tmat = tmat[selected_cells,]
distHC <- c("distPearson","distCosine","euclidean","maximum","manhattan","canberra","binary","minkowski")[3]
methHC <- c("ward","ward.D","ward.D2","single","complete","average")[2]
if(distHC=="distPearson")	d <- distPearson(tmat)
if(distHC=="distCosine")	d <- distCosine(tmat)
if(distHC %in% c("euclidean","maximum","manhattan","canberra","binary","minkowski"))	d <- dist(tmat,distHC)
hc <- hclust(d,method=methHC)
mat.so <- pathway_mat[,hc$order]
 
if(chRangeHM){
	for(i in 1:nrow(mat.so)){
		mat.so[i,] <- geco.changeRange(mat.so[i,],newmin=0,newmax=1)
	}
}
rowClust <- hclust(as.dist(1 - cor(t(mat.so))), method = "ward.D")

png(file.path(resSUBdir,paste0("Clustering_Pathways_Persisters_homemade_",distHC,"_",methHC,".png")), height=1000,width=1000,res=150)
geco.hclustAnnotHeatmapPlot(x=(mat.so[rowClust$order,]),
							 hc=hc,
							 hmColors=hmColors,
							 anocol=anocol2[hc$order,c(1,8)],#[,ncol(cc.col):1]
							 xpos=c(0.15,0.9,0.164,0.885),
							 ypos=c(0.1,0.5,0.5,0.6,0.62,0.95),
							 dendro.cex=0.1,
							 xlab.cex=0.6,
							 hmRowNames=TRUE,
							 hmRowNames.cex=0.5
							 )
dev.off()
nclust <- 5
annot_sel2$clust_heatmap <- paste("C",match(cutree(hc,nclust),unique(cutree(hc,nclust)[hc$order])),sep="")
pdf(file.path(resSUBdir,"Clustering_Pathways_Persisters.pdf"))
heatmap.2(pathway_mat,ColSideColors=anocol2[,1],trace="none")
dev.off()

```
